Museum specimens as baseline for assessing a potential global pollinator crisis

### 2018-03-15 ms.Rmd
ms edited again

There is ever-increasing concern about the decline of pollinators, especially for bees. However, despite reports that pollinator declines are widespread, data is scarce and geographically and taxonomically biased. This limits any robust inference about a potential pollination crisis. The most direct approach for identifying patterns in species declines is analysis of historical time series data. However, these data are often rare or non-existent for most pollinator groups, but provide the only source of historical information serving as a baseline to understand the extent of pollinator declines. Specimens historically collected by naturalists and preserved in museums can provide information on when and where species were collected, as well as other ecological information (e.g., historical species interactions and morphological traits). Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current occurrence patterns to identify changes in biodiversity, relative species abundances, body size and even pollination services over time. Despite recent advances, we show that information on the status and trends of most pollinators is missing and a key bottle neck slowing progress in this field is the digitization of historical data. We identify opportunities and limitations to progress understanding of pollinator declines worldwide. Further, we demonstrate how to analyse historical museum collection data, by providing two contrasting case studies from distinct geographical regions, New Zealand and Spain, for which historical pollinator declines have never been assessed. We show that there is a great potential for museum specimens to play a key role in identifying the extent of the pollination crisis if focus on unleashing the power of historical collection data.

@@

Animal pollinators are acritical component of ecosystems, given their role in plant reproduction (@ollerton2012) and food security worldwide (@kleinetal2007). As many other taxa, pollinators are vulnerable to multiple human-mediated environmental changes, which potentially threatening their populations. This problem was identified by researchers several decades ago, and was popularized by the influential book "the forgotten pollinators" (1996). However, these accounts were relatively anecdotal as there were almost no data on pollinator trends and statuses at the time. These initial claims triggered the first efforts to address the problem, including a US National Academy of Science (NAS) panel commissioned to assess the extent of pollinators declines (2006). The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult". Since then, studies on pollinator responses to global change have multiplied and researchers have reached strong consensus that pressures such as habitat destruction, land use change, chemical exposure, exotic species and climate change are the primary causes of pollinator declines, and often act synergistically (@goulson2016, @potts2010). Yet the status and trends of most pollinator species worldwide remain unknown. For example, even for a well-studied region such as Europe, 60% [check number]() of bee species assessed by in recent IUCN report fall into the category "Data Deficient" (@Nieto2014). For countries outside of Europe and the US, data on pollinator population is almost non-existent.

One of the main barriers to identifying pollinator population trends is that pollinators are a diverse group of taxa including bees, flies, butterflies and beetles, and also birds, bats and even lizards (@Winfreeetal2011). In particular, bees are usually regarded as the more important pollinaotrs due to their abundance, pollination efficiency and widespread distribution (@ollerton2018). However, bees are an incredibly diverse group, with more than 20,000 species worldwide (@PickeringAscher), and most require taxonomic expertise for identification. This has led to a biased geographic patterns in bee decline research due to the uneven distribution of researchers (@Martinetal2012), but also taxonomic biases toward species that are easier to identify, such bumblebees (@ArcherOikos). Another barrier is methodological and involves challenges for assessing population trends for highly mobile organisms, which are also often short-lived and small.
One solution to overcoming these barriers has been to use space for time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach (@ref), this is the most extensive source of data currently available and in Europe, researchers have recently estimated bee richness declines for each country using predictions generated from models (@DePalma). A second promising method is pollinator monitoring programs, with data often collected by citizen scientists. This approach was inspired by successful butterfly monitoring programs and is currently being extended to other pollinator taxa (@ref). The main limitation is that these programs need significant time to generate long-term datasets and do not reveal anything about historic pollinator communities. Finally, the most robust approach for identifying historical pollinator tends is to use museum collection data. The core aim of museums is precisely to conserve and curate collections, thus they not only serve as a precious repository specimens, but often also ensure high quality taxonomic identification (@primak). However, the major bottleneck for researchers wanting to use this information is the digitization of specimens. Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have facilitated identification of current status and trends in pollinator populations (e.g. Bartomeus et al 2013, Scheper et al 2014). This is despite the fact that museum collections often have a number of biases (@Keil). In reality, museum collections are the only data we have to reconstruct the past.

Regardless of the method, it is important to consider that biodiversity has many aspects beyond species richness, and we should into identifying changes not only in local (alpha) diversity or regional (gamma) diversity, but also in turnover between sites (beta diversity), abundance of key species, and evenness of communities (@winfree2016ELE). Environmental changes often results in few winners and many of losers (@Bartomeusetal2013). Identifying winners and losers is critical as the few winners are often exotic species, and represent a subset of trait that facilitate survival in highly modified environments (@Stavertetal2018). These changes can have important flow on effects for pollination of native species and crops.  [Add something about traits of winners and losers?]()
https://www.sciencedirect.com/science/article/pii/S0065250417300296
[move this and expand to the discussion on the map?]()

@@

Digitizing old specimens is not a trivial task and it needs expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geo-locate the coordinates of the collection event (@ref and software) and (iii) store the data in a properly curated database (@ref on database management). Undertaking this process for the thousands of specimens a museum collection is a daunting task and requires specialized personnel. While some tasks can only be undertaken by those with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the process of digitizing collections. High resolution photos of specimens and the associated labels can be uploaded to the internet, where the work of transcribing the images can be distributed across hundreds or thousands of volunteers. Furthermore, new algorithms have been created to geo-reference localities based on vernacular names (@refs geonames, geoparser and opencage). Explain zooniverse, NZ initiative and how data can be validated... [Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]
[how much do we want to extend on this? Is interesting info, but boring o read?]()

@@

As illustrated in this examples, unleashing the power of museum data to answer pressing questions is at our hands, and need the coordinated effort of many actors. As evidenceed by the case studies, collaborating with museum curators and original data collectors is key to understand the data and treat it appropiatelly. We envision that rather than a central repository of museum records, we should aim to make the digitalized data available and easy to share, and if possible ultimately centralized in existing platforms like Gbif. Even when the data lives in different museum webpages or databases, if it's easily accessible and has programmatic access (API e.g. @heltmint) it can be retried and joined with other data sets using open science tools (@ropensci).

We also need to revalorize museums and foster a birectional relationship between researchers and museums. Researchers collect massive amounts of specimens, not always appropriately vouched. Depositing this in museums should ensure its long term preservation. However, this require appropriately funded museums and recognizing data (and not only papers) as an important research output.

[What about an interactive bee decline map? or Integration of DATA and analysis:]()

Ultimately, we need more data to compare global differences geographically and see which are the generalized trends. As far as we know there have been few bee regional extinctions (@Ollerton, @Bartomeus2013), but bee declines are widespread in disturbed ecosystems of north america and Europe (@Bartomeus, @dePalma, @Carvalheiro). We know almost nothing about the rest... for those areas, the first step is to know what pollinators exists and name the species before we loss them.

@@

E. Asensio (or maybe coauthor?). Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum, ITACyL. Barry Donovan for providing New Zealand bee collection records. Carola Warner for entering the New Zealand data.

>>>>

There is increasing concern about the decline of pollinators worldwide. However, despite reports that pollinator declines are widespread, data is scarce and often geographically and taxonomically biased. This limits robust inference about any potential “pollination crisis”. [something about how to address the problem]. Analysis of historical time series data is the most direct approach for assessing species declines. In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines. Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits). However, data of this kind are typically rare or non-existent for most pollinator taxa. Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and even pollination services. Despite recent advances, we show that information on the status and trends of most pollinators is absent, and that the lack of collection record data digitization is a key bottleneck, impeding progress in this field. We identify opportunities and limitations to progress in assessing pollinator declines at a global scale. Further, we demonstrate various approaches to analysing museum collection data by providing two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed. Finally, we suggest that there is immense potential for museum specimens to play a central role in assessing the extent of the pollination crisis globally.

@@

Animal pollinators are a critical component of both natural and production ecosystems worldwide, given their role in plant reproduction (@ollerton2012) and food security (@kleinetal2007). As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. This issue was identified several decades ago, and was popularized in 1996 by the influential book "the forgotten pollinators". However, these early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time. These initial claims triggered the first efforts to asses the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult". Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically (@goulson2016, @potts2010). Yet the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category (@Nieto2014). For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and including bees, flies, butterflies, beetles, birds, bats and even lizards (@Winfreeetal2011). Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution (@ollerton2018). However, bees are also extremely diverse, with more than 20,000 species currently described worldwide (@PickeringAscher), and require often require expert taxonomists for identification. The uneven distribution of bee researchers has resulted in geographical biases in bee decline research (@Martinetal2012) and there are also taxonomic biases toward species that are easier to identify, such bumblebees (@ArcherOikos). Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.
One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach (@ref), this is the most extensive source of data on pollinator populations currently available. In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models [using what parameters?] (@DePalma). A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs in countries such as the UK and is currently being extended to other pollinator taxa (@ref). However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations. Finally, the most robust approach for assessing long-term pollinator population trends is to use museum specimen collection data. The core aim of museums is to conserve and curate collections. Thus, they serve as a precious repository for specimens and also ensure high quality taxonomic identification (@primak). Yet currently, the major bottleneck for researchers wanting to use contained within museum collections is the digitization of collection data. Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g. Bartomeus et al 2013, Scheper et al 2014). This is despite the fact that museum collections often have a number of biases [including...] (@Keil). But in many cases, museum collections are the only data we have to reconstruct the past.

Regardless of the method, it is important for researchers to consider aspects of biodiversity beyond species richness. In addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities (@winfree2016ELE). Environmental changes often results in few “winners” and many “losers” (@Bartomeusetal2013). Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments (@Stavertetal2018). These changes can have important flow on effects for pollination of native plant species and crops.  [Add something about traits of winners and losers?]()
https://www.sciencedirect.com/science/article/pii/S0065250417300296
[move this and expand to the discussion on the map?]()

@@

Digitizing old specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geo-locate the coordinates of collection events (@ref and software) and (iii) store the data in a properly curated database (@ref on database management). Undertaking this process for the thousands of specimens in museum collections can be a daunting task and requires specialized personnel. While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process. High resolution specimen photos and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers. In addition to this, new algorithms have been created that allow location geo-referencing based on vernacular names (@refs geonames, geoparser and opencage). [Explain zooniverse, NZ initiative and how data can be validated... Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]

[how much do we want to extend on this? Is interesting info, but boring o read?]()

@@

As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors. We show in our case studies that collaboration with both museum curators and data collectors is key to understanding the data and treating it appropriately. We envision that rather than a central repository of museum collection records, we should aim to make digitalized data readily available and easy to share. Ultimately, centralisation of these data in existing platforms such as Gbif would facilitate free and widespread access. However, even when data is stored in different webpages or databases, it can be retried and joined with other data sets using open science tools (@ropensci) if it is easily accessible and has programmatic access (API e.g. @heltmint).

We also need to revolutionize the way that researchers collaborate with museums by foresting healthy bidirectional relationships. For instance, researchers collect massive numbers of specimens, but these are often inappropriately vouched, rendering them useless for research purposes. To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is adequate funding and recognition that accurate data recording and long term preservation are critical for research.
[What about an interactive bee decline map? or Integration of DATA and analysis:]()

Ultimately, to identify global trends in pollinator declines we require data from diverse geographic regions. In particular, we need to put increased emphasis on quantifying pollinator declines in regions outside of the US and Europe. Furthermore, we need to investigate whether there have been declines in pollinator groups other than bees. As far as we know, there have been few regional bee extinctions (@Ollerton, @Bartomeus2013), but declines are widespread in disturbed ecosystems of the US and Europe (@Bartomeus, @dePalma, @Carvalheiro). For most other pollinator taxa we know almost nothing. For many taxa, the first step is to identify and describe them and only then can we begin to document declines.

@@

We thank E. Asensio (or maybe coauthor?), Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum, ITACyL. We also thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.
